---
title: "Coming from grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coming from grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
```

vellum sits at grid's level of the stack: units, viewports, grobs, layout, and
rendering. If you know grid, most concepts carry over directly. This guide maps
the vocabulary and then flags the handful of places where vellum works
differently on purpose.

## The concept map

| grid | vellum | notes |
|------|--------|-------|
| `grid.newpage()` | `vl_scene()` | vellum's scene also fixes page size, background, and dpi up front |
| `unit(1, "native")` | `unit(1, "native")` | same idea; each element carries its own unit |
| `viewport()` | `viewport()` | region with its own `xscale` / `yscale` |
| `pushViewport()` / `popViewport()` | `push()` / `pop()` | functional: they take and return a scene, no global stack |
| `grid.rect()`, `grid.circle()`, ... | `rect_grob()`, `circle_grob()`, ... plus `draw()` | the constructor builds a value; `draw()` adds it |
| `rectGrob()`, `gTree()` | `rect_grob()`, the scene tree | grobs are immutable S7 values |
| `gpar()` | `gpar()` | familiar fields; `fill` also accepts gradients |
| `grid.layout()` | `grid_layout()` | flexible `"null"` tracks work the same |
| `grid.edit()` / `editGrob()` | `edit_node()` | edit by `name`; copy-on-modify, not in place |
| `grid.grabExpr()` / display list | the retained scene | the tree is the model; nothing is replayed |
| `grid.locator()` | `hit_test()` | exact geometric picking, not one interactive click |
| device (`png()`, `pdf()`, ...) | `render(scene, path)` | the extension picks the backend |

## Side by side

A minimal grid plot and its vellum translation. In grid:

```r
library(grid)
grid.newpage()
pushViewport(viewport(width = 0.8, height = 0.8,
                      xscale = c(0, 10), yscale = c(0, 20)))
grid.rect(gp = gpar(fill = "grey97", col = "grey70"))
grid.lines(x = unit(0:10, "native"), y = unit((0:10) * 2, "native"),
           gp = gpar(col = "steelblue", lwd = 2))
popViewport()
```

The same scene in vellum:

```{r}
#| label: translation
vl_scene(5, 3, bg = "white") |>
  push(viewport(width = 0.8, height = 0.8,
                xscale = c(0, 10), yscale = c(0, 20))) |>
  draw(rect_grob(gp = gpar(fill = "grey97", col = "grey70"))) |>
  draw(lines_grob(x = unit(0:10, "native"), y = unit((0:10) * 2, "native"),
                  gp = gpar(col = "steelblue", lwd = 2))) |>
  pop()
```

The shapes are the same; the difference is that the vellum version is one
expression that returns a scene value, with no global device or viewport stack
mutated along the way.

## What is different, and why

### The builder is functional, not stateful

grid keeps a global viewport stack and a display list: `pushViewport()` mutates
state, and each `grid.*` call paints into the current device. vellum's `push()`,
`draw()`, and `pop()` each take a scene and return a new one. There is no
"current viewport" to lose track of, the pipe *is* the tree, and a scene is an
ordinary value you can store, pass around, and branch from.

### Metrics are eager, so there is no draw-time hook protocol

grid cannot know a grob's size until a device and viewport exist at draw time,
which is why it has lazy units and the `makeContext` / `makeContent` /
`widthDetails` hook protocol, and why it replays the whole display list on
resize. vellum resolves text and object metrics in process, up front, so a grob
knows its extent when you build it. You measure with `vl_strwidth()` or size a
unit by a grob's extent with `grobwidth()` / `grobheight()` immediately, without
an open device.

### Units resolve eagerly, and mixed-unit arithmetic is restricted

Because units resolve eagerly to a flat representation, vellum will *not* defer
`unit(1, "npc") - unit(2, "mm")` the way grid does. Same-space arithmetic and
absolute-plus-absolute both work (`unit(10, "mm") + unit(1, "in")` gives
`35.4mm`), but mixing a relative and an absolute unit in one expression is
reported rather than guessed.

```{r}
#| label: unit-arith
unit(10, "mm") + unit(1, "in") # absolute + absolute -> mm
unit(1:3, "cm") * 2            # scaling is fine
```

If you have grid code that offsets a relative position by an absolute amount,
compose it at the viewport or native level instead of in a single `unit`
expression. This is the change most likely to surface when porting grid code.

### The tree is retained and inspectable

grid's rendered output is pixels plus a display list. vellum keeps the scene as
an immutable tree you can query and edit after the fact: `node_names()`,
`get_node()`, `edit_node()`, `hit_test()`, and `scene_model()`. There is no
`grid.force()` step, and editing a node copies rather than mutating in place.
See `vignette("retained-mode")` for what this enables.

### `gpar` inherits, but there is no cascade

`gpar()` inheritance works as in grid: a field left `NULL` is inherited from the
enclosing viewport, and `alpha` multiplies down the tree. vellum does not add a
CSS-like cascade with selectors or theme objects at this layer; that is a
grammar-layer concern.

## Do I have to rewrite my grid code?

Not necessarily. If you already have grid, ggplot2, or lattice output, you can
render it *through* vellum's backend without porting anything, using
`as_vellum()` and `render_grid()`. See `vignette("grid-interop")`. Rewriting in
the native API is worth it when you want the retained-tree features (naming,
editing, hit-testing) or deterministic multi-backend output.
```
