---
title: "Coming from grid"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Coming from grid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
```

vellum sits at grid's level of the stack: units, viewports, grobs, layout, and
rendering. If you know grid, most concepts carry over directly. This guide maps
the vocabulary and then flags the handful of places where vellum works
differently on purpose.

## The concept map

| grid | vellum | notes |
|------|--------|-------|
| `grid.newpage()` | `vl_scene()` | vellum's scene also fixes page size, background, and dpi up front |
| `unit(1, "native")` | `vl_unit(1, "native")` | same idea; each element carries its own unit |
| `viewport()` | `vl_viewport()` | region with its own `xscale` / `yscale` |
| `pushViewport()` / `popViewport()` | `push()` / `pop()` | functional: they take and return a scene, no global stack |
| `grid.rect()`, `grid.circle()`, ... | `rect_grob()`, `circle_grob()`, ... plus `draw()` | the constructor builds a value; `draw()` adds it |
| `rectGrob()`, `gTree()` | `rect_grob()`, the scene tree | grobs are immutable S7 values |
| `gpar()` | `vl_gpar()` | familiar fields; `fill` also accepts gradients |
| `grid.layout()` | `grid_layout()` | flexible `"null"` tracks work the same |
| `grid.ls()` / `grid.get()` | `node_names()` / `get_node()` | list and read nodes by `name` |
| `grid.edit()` / `editGrob()` | `edit_node()` | `editGrob()` also copies; `grid.edit()` mutates the display list, `edit_node()` never does |
| `grid.grabExpr()` / display list | the retained scene | the tree is the model, not something you recover from a device |
| `grid.locator()` | `hit_test()` | picking from a pick-buffer, not one interactive click |
| (no counterpart) | `scene_model()` | per-element identity plus resolved device-pixel bbox |
| device (`png()`, `pdf()`, ...) | `render(scene, path)` | the extension picks the backend |

## Side by side

A minimal grid plot and its vellum translation. In grid:

```r
library(grid)
grid.newpage()
pushViewport(viewport(width = 0.8, height = 0.8,
                      xscale = c(0, 10), yscale = c(0, 20)))
grid.rect(gp = gpar(fill = "grey97", col = "grey70"))
grid.lines(x = unit(0:10, "native"), y = unit((0:10) * 2, "native"),
           gp = gpar(col = "steelblue", lwd = 2))
popViewport()
```

The same scene in vellum:

```{r}
#| label: translation
vl_scene(5, 3, bg = "white") |>
  push(vl_viewport(width = 0.8, height = 0.8,
                xscale = c(0, 10), yscale = c(0, 20))) |>
  draw(rect_grob(gp = vl_gpar(fill = "grey97", col = "grey70"))) |>
  draw(lines_grob(x = vl_unit(0:10, "native"), y = vl_unit((0:10) * 2, "native"),
                  gp = vl_gpar(col = "steelblue", lwd = 2))) |>
  pop()
```

The shapes are the same; the difference is that the vellum version is one
expression that returns a scene value, with no global device or viewport stack
mutated along the way.

## What is different, and why

### The builder is functional, not stateful

grid keeps a global viewport stack and a display list: `pushViewport()` mutates
state, and each `grid.*` call paints into the current device. vellum's `push()`,
`draw()`, and `pop()` each take a scene and return a new one. There is no
"current viewport" to lose track of, the pipe *is* the tree, and a scene is an
ordinary value you can store, pass around, and branch from.

### Metrics are eager, so there is no draw-time hook protocol

grid cannot know a grob's size until a device and viewport exist at draw time,
which is why it has lazy units and the `makeContext` / `makeContent` /
`widthDetails` hook protocol, and why it replays the whole display list on
resize. vellum resolves text and object metrics in process, up front, so a grob
knows its extent when you build it. You measure with `vl_strwidth()` or size a
unit by a grob's extent with `grobwidth()` / `grobheight()` immediately, without
an open device.

### Units are records, not expression trees

A grid unit defers its arithmetic: `unit(1, "npc") - unit(2, "mm")` is stored as
`sum(1npc, -2mm)` and evaluated when a device and viewport exist. A vellum unit is
a flat record — a value, a base code, and an absolute millimetre offset — so
arithmetic reduces at construction wherever it can.

```{r}
#| label: unit-arith
vl_unit(10, "mm") + vl_unit(1, "in")     # absolute + absolute -> mm
vl_unit(1, "native") + vl_unit(2, "mm")  # a compound unit: base + offset
vl_unit(1:3, "cm") * 2                   # scaling is fine
```

The third form is the one worth knowing. A position base plus an absolute is kept
as a *compound* unit that resolves as "the data (or npc) position, displaced by
exactly this many millimetres" at any scale, aspect, or dpi — which is the case
most grid code that mixes units actually wants.

The one thing grid expresses and vellum does not is a mix of two *different*
position bases (`"npc"` plus `"native"`), because that cannot reduce to a single
record without a device. It errors, with a message naming the legal forms. Put the
normalised part in the viewport and the data part in the coordinate, or pre-resolve
to absolute units.

### The tree is a value, and it can report geometry

grid retains its scene as well, so this is a difference of degree for the
inspect-and-edit half: `grid.ls()`, `grid.get()`, `grid.edit()` and `editGrob()`
already let you rewrite a plot after the fact, and `node_names()`, `get_node()`
and `edit_node()` are the counterparts here. The difference is that vellum's tree
is an immutable value rather than device state, so there is no `grid.force()`
step and no in-place mutation of a display list.

The half with no grid counterpart is geometry read-back: `hit_test()` picks the
topmost node under a point, and `scene_model()` returns every drawn element with
its resolved device-pixel bounding box. See `vignette("retained-mode")` for what
that enables.

### `vl_gpar` inherits, but there is no cascade

`vl_gpar()` inheritance works as in grid: a field left `NULL` is inherited from the
enclosing viewport, and `alpha` multiplies down the tree. vellum does not add a
CSS-like cascade with selectors or theme objects at this layer; that is a
grammar-layer concern.

## Do I have to rewrite my grid code?

Not necessarily. If you already have grid, ggplot2, or lattice output, you can
render it *through* vellum's backend without porting anything, using
`as_vellum()` and `render_grid()`. See `vignette("grid-interop")`. Rewriting in
the native API is worth it when you want the retained-tree features (naming,
editing, hit-testing) or deterministic multi-backend output.
```
