---
title: "Rendering grid and ggplot2 through vellum"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering grid and ggplot2 through vellum}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96,
  dev = "png"
)
library(vellum)
```

You do not have to rewrite existing graphics to draw them with vellum's backend.
`as_vellum()` converts a grid grob tree, or a ggplot2 plot, into a `vl_scene()`,
and `render_grid()` does that and writes the result to a file. This is the
interop path: grid-based graphics rendered by vellum's deterministic PNG / SVG /
PDF backend.

## How it works

grid graphics are lazy: a grob's real geometry is only known once a device and
viewport resolve its units. `as_vellum()` leans on that machinery rather than
reimplementing it. It spins up an offscreen grid device, lets grid resolve every
coordinate to an absolute position, and then emits the corresponding vellum
grobs. The output is a normal `vellum_scene`, so everything else in vellum
(multi-backend `render()`, `scene_raster()`, `display()`) applies.

## A grid grob tree

Any grob or gTree works. Here is a small hand-built one.

```{r}
#| label: grid-grob
library(grid)

g <- gTree(children = gList(
  rectGrob(gp = gpar(fill = "grey97", col = "grey70")),
  circleGrob(r = 0.3, gp = gpar(fill = "#3a7bd5", col = NA)),
  textGrob("grid -> vellum", y = 0.12, gp = gpar(fontface = "bold"))
))

as_vellum(g, width = 5, height = 3)
```

`as_vellum()` returns a scene, which auto-prints (displays) here. Assign it and
you can render it to any backend:

```{r}
#| label: render-grid
#| eval: false
render_grid(g, "grid.png", width = 5, height = 3)
render_grid(g, "grid.pdf", width = 5, height = 3)
```

## A ggplot2 plot

Passing a ggplot object works the same way: ggplot builds a gtable of grobs, and
`as_vellum()` renders that.

```{r}
#| label: ggplot
#| eval: !expr requireNamespace("ggplot2", quietly = TRUE)
library(ggplot2)

p <- ggplot(mtcars, aes(wt, mpg, colour = factor(cyl))) +
  geom_point(size = 2) +
  labs(title = "mtcars", colour = "cyl") +
  theme_minimal()

as_vellum(p, width = 6, height = 4)
```

To write it out, use `render_grid()` with the format in the file extension:

```{r}
#| label: ggplot-render
#| eval: false
render_grid(p, "mtcars.png", width = 6, height = 4)
render_grid(p, "mtcars.svg", width = 6, height = 4)
```

lattice output works through the same door: draw the lattice object to capture
its grob tree, or pass a captured grob to `as_vellum()`.

## When to use interop versus the native API

Reach for `as_vellum()` / `render_grid()` when you already have grid, ggplot2,
or lattice output and simply want vellum's deterministic, multi-backend
rendering (for example byte-stable PNGs for snapshot tests, or one plot emitted
to PNG, SVG, and PDF from the same source).

Build with the native `vl_scene()` API instead when you want what the retained
scene graph offers: named and editable nodes, hit-testing, and a per-element
`scene_model()` (see `vignette("retained-mode")`). Interop faithfully reproduces
the *pixels* of a grid scene, but a grammar built directly on vellum is what
carries the per-element identity that interactivity needs.

A native vellum graphics device (so `plot()` and friends target vellum directly)
is future work; until then, `as_vellum()` is the bridge.
```
