---
title: "Inspecting a scene: linting, coverage, profiling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inspecting a scene: linting, coverage, profiling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
set.seed(1)
```

A vellum scene is not only something you render. Because layout and text metrics
are resolved **before** anything is drawn, a finished scene can be asked
questions — where will this land, how big will that text be, what is expensive —
and answer them exactly, without opening a device.

This article covers the three tools built on that: `vl_lint()` finds likely
mistakes, `scene_stats()` measures coverage and crowding, and
`profile_render()` says where the time goes.

## Why this can exist here

In grid, a grob's size and position are not knowable until draw time: text has
no measurable width before a device is open, `native` units need a viewport, and
`null` units need a layout context. Anything wanting to check "is this label
readable at the rendered size" would have to re-implement the geometry and hope
it stays in step with the renderer.

vellum resolves all of it up front, and every tool below reads *the same
resolved numbers the renderer draws with*. That is the whole trick.

## `vl_lint()`: static analysis for graphics

Here is a plot with four defects planted in it. Each is the kind of thing you
normally discover by exporting the figure and squinting at it.

```{r}
bad <- vl_scene(5, 3, dpi = 96) |>
  push(vl_viewport(name = "panel", y = 0.55, width = 0.85, height = 0.7,
                   xscale = c(0, 1), yscale = c(0, 1), clip = TRUE)) |>
  draw(rect_grob(gp = vl_gpar(fill = "grey96", col = "grey70"))) |>
  draw(points_grob(runif(120), runif(120), size = vl_unit(1.6, "mm"),
                   gp = vl_gpar(fill = "#7FB2E5", col = NA), name = "cloud")) |>
  # 1. outside the panel's clip: silently never drawn
  draw(points_grob(2.4, 0.5, size = vl_unit(3, "mm"),
                   gp = vl_gpar(fill = "tomato", col = NA), name = "stray_point")) |>
  # 2. too small to read
  draw(text_grob("n = 120", x = 0.06, y = 0.93, just = c("left", "top"),
                 gp = vl_gpar(fontsize = 2.5), name = "n_label")) |>
  # 3. will not contrast with the panel behind it
  draw(text_grob("watermark", gp = vl_gpar(fontsize = 22, col = "#F2F2F2"),
                 name = "watermark")) |>
  # 4. nothing will ever paint it
  draw(rect_grob(x = 0.8, y = 0.2, width = 0.2, height = 0.1,
                 gp = vl_gpar(fill = NA, col = NA), name = "empty_box")) |>
  pop()

vl_lint(bad)
```

Every finding names the grob, so you can go straight to it. Nothing here needed
you to look at the picture.

The rules that ship:

```{r}
vl_lint_rules()
```

Two are worth explaining.

`tiny_text` measures in **device pixels at the size you are actually
rendering**, not in points. A 6 pt label is fine in a print figure and illegible
in a thumbnail, and the same scene rendered at two sizes should — correctly —
lint differently. Adjust the floor with `min_text_px`.

`low_contrast` samples the **rendered image** just outside each label's box, on
four sides, and takes the worst. So it measures the backdrop the label really
sits on, including any marks that happen to be behind it, rather than assuming
the page colour:

```{r}
faint <- vl_scene(4, 1.2, dpi = 96, bg = "white") |>
  draw(text_grob("hard to read", gp = vl_gpar(fontsize = 22, col = "#DDDDDD")))
vl_lint(faint)

# Same text colour, dark page: now it is fine, and the linter agrees.
vl_lint(vl_scene(4, 1.2, dpi = 96, bg = "grey15") |>
  draw(text_grob("easy to read", gp = vl_gpar(fontsize = 22, col = "#DDDDDD"))))
```

### Adding your own rules

The rule set is a registry, which is the point: vellum knows about geometry, but
it knows nothing about *your* semantics. A package layered on top can add rules
that do, and both come back from one `vl_lint()` call.

A rule is a function of `(scene, nodes, ctx)`. `nodes` is the resolved per-node
table — device-px box, clip box, kind, name, element count, alpha, whether a
fill or stroke is present, font size in px, colour, label — and `ctx` carries the
page size, dpi, thresholds, and a `pixel(x, y)` sampler.

```{r}
vl_lint_rule(
  "wide_marks",
  function(scene, nodes, ctx) {
    wide <- nodes[(nodes$x1 - nodes$x0) > 0.8 * ctx$w, , drop = FALSE]
    vl_lint_finding("wide_marks", "note", wide, "spans almost the whole page")
  },
  description = "A mark spans nearly the full page width."
)

vl_lint(bad, rules = c("wide_marks", "tiny_text"))
```

Register from your package's `.onLoad()` and your users get your rules for free.

## `scene_stats()`: how crowded is this, really?

A mark count tells you nothing about whether the marks land on top of each
other. `scene_stats()` measures what actually reached the canvas:

```{r}
cloud <- function(n, sd = NA) {
  xy <- if (is.na(sd)) list(runif(n), runif(n)) else list(rnorm(n, 0.5, sd), rnorm(n, 0.5, sd))
  vl_scene(4, 3, dpi = 96) |>
    draw(points_grob(xy[[1]], xy[[2]], gp = vl_gpar(fill = "steelblue", col = NA)))
}

round(rbind(
  scattered = scene_stats(cloud(2000)),
  clustered = scene_stats(cloud(2000, sd = 0.03))
), 3)
```

Identical element counts; wildly different plots. `overplot` estimates how many
times an average inked pixel was painted over — `1` means no overlap — and it is
the honest trigger for reaching for `datashade()`. Thirty thousand well-separated
points are fine; eight thousand piled up are not, and only this number knows the
difference.

It is computed from bounding boxes, so treat it as an index rather than a
measurement: a circle fills about 79% of its box, and text and unkeyed
lines/polygons contribute to `ink` but not to `overplot`.

## `profile_render()`: where the time goes

```{r}
heavy <- vl_scene(6, 4, dpi = 96) |>
  draw(points_grob(runif(20000), runif(20000), name = "points")) |>
  draw(segments_grob(runif(2000), runif(2000), runif(2000), runif(2000), name = "edges")) |>
  draw(text_grob(format(1:200), x = runif(200), y = runif(200),
                 gp = vl_gpar(fontsize = 8), name = "labels"))

profile_render(heavy, reps = 1)
```

**Read the phase split first.** A render is three phases: *build* (constructing
the R value), *compile* (the R→Rust replay, including text shaping), and *raster*
(drawing). If compile dominates — which it does on scenes made of many small
grobs, where the cost is S7 validation and grob construction — then the backend
is not your problem and tuning it will not help. `profile_render()` says so
explicitly when that happens.

The per-mark table covers raster time, aggregated by mark rather than by node: a
vectorised `text_grob()` of 200 labels compiles to 200 nodes, and 200 rows of
almost nothing is noise rather than a profile.

Timing is armed only for that call, so ordinary renders pay nothing for it.

## Putting it in a test

All three return plain data frames, so they drop straight into a test suite. A
lint check is a reasonable thing to assert in CI:

```{r}
#| eval: false
test_that("the figure has no lint findings", {
  expect_equal(nrow(vl_lint(my_plot())), 0L)
})
```

## Where to go next

- `vignette("render-quality")`: colour-vision simulation, crisp gridlines, and
  group effects — the other half of "will this actually read".
- `vignette("retained-mode")`: `scene_model()`, `hit_test()`, and editing nodes
  by name.
