---
title: "Inspecting a scene: linting, coverage, profiling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inspecting a scene: linting, coverage, profiling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
set.seed(1)
```

A vellum scene is not only something you render. Because layout and text metrics
are resolved **before** anything is drawn, a finished scene can be asked
questions — where will this land, how big will that text be, what is expensive —
and answer them exactly, without opening a device.

This article covers the three tools built on that: `vl_lint()` finds likely
mistakes, `scene_stats()` measures coverage and crowding, and
`profile_render()` says where the time goes.

## Why this can exist here

In grid, a grob's size and position are not knowable until draw time: text has
no measurable width before a device is open, `native` units need a viewport, and
`null` units need a layout context. Anything wanting to check "is this label
readable at the rendered size" would have to re-implement the geometry and hope
it stays in step with the renderer.

vellum resolves all of it up front, and every tool below reads *the same
resolved numbers the renderer draws with*. That is the whole trick.

## `vl_lint()`: static analysis for graphics

Here is a plot with four defects planted in it. Each is the kind of thing you
normally discover by exporting the figure and squinting at it.

```{r}
bad <- vl_scene(5, 3, dpi = 96) |>
  push(vl_viewport(name = "panel", y = 0.55, width = 0.85, height = 0.7,
                   xscale = c(0, 1), yscale = c(0, 1), clip = TRUE)) |>
  draw(rect_grob(gp = vl_gpar(fill = "grey96", col = "grey70"))) |>
  draw(points_grob(runif(120), runif(120), size = vl_unit(1.6, "mm"),
                   gp = vl_gpar(fill = "#7FB2E5", col = NA), name = "cloud")) |>
  # 1. outside the panel's clip: silently never drawn
  draw(points_grob(2.4, 0.5, size = vl_unit(3, "mm"),
                   gp = vl_gpar(fill = "tomato", col = NA), name = "stray_point")) |>
  # 2. too small to read
  draw(text_grob("n = 120", x = 0.06, y = 0.93, just = c("left", "top"),
                 gp = vl_gpar(fontsize = 2.5), name = "n_label")) |>
  # 3. will not contrast with the panel behind it
  draw(text_grob("watermark", gp = vl_gpar(fontsize = 22, col = "#F2F2F2"),
                 name = "watermark")) |>
  # 4. nothing will ever paint it
  draw(rect_grob(x = 0.8, y = 0.2, width = 0.2, height = 0.1,
                 gp = vl_gpar(fill = NA, col = NA), name = "empty_box")) |>
  pop()

vl_lint(bad)
```

Every finding names the grob, so you can go straight to it. Nothing here needed
you to look at the picture.

### Seeing the findings

For a graphics linter, a message describing a defect is second best. `vl_lint_overlay()`
draws the report onto the scene — a box round each finding, labelled with the
rules that fired, red for warnings and orange for notes:

```{r}
#| fig.alt: "The scene above with red and orange boxes drawn around each linted mark."
display(vl_lint_overlay(bad))
```

The box around `empty_box` is the one to notice: there is nothing there to see,
which is precisely the finding.

### The rules that ship

```{r}
vl_lint_rules()[, c("rule", "description")]
```

Findings come at two severities. A **warning** is a probable defect — text
nobody can read, a mark that never reaches the page. A **note** is something to
look at, because the same geometry is sometimes deliberate: a cropped mark, a
label inside the bar it belongs to.

A few rules are worth explaining.

**`tiny_text` has two floors, and fires on either.** Device pixels decide
whether glyphs survive rasterisation; points decide whether a human can read
them. A pixel floor alone goes blind at print resolution, because `font_px`
scales with `dpi` — 4 pt at 300 dpi is 16.7 px, comfortably clear of the 7 px
default and still unreadable on paper:

```{r}
tiny <- function(dpi) {
  vl_scene(3, 1, dpi = dpi) |>
    draw(text_grob("4 pt caption", gp = vl_gpar(fontsize = 4)))
}
vl_lint(tiny(96))$message
vl_lint(tiny(300))$message
```

**`low_contrast` samples the rendered image**, just outside each label's box on
four sides, so it measures the backdrop the label really sits on — including
marks behind it — rather than assuming the page colour:

```{r}
faint <- vl_scene(4, 1.2, dpi = 96, bg = "white") |>
  draw(text_grob("hard to read", gp = vl_gpar(fontsize = 22, col = "#DDDDDD")))
vl_lint(faint)

# Same text colour, dark page: now it is fine, and the linter agrees.
vl_lint(vl_scene(4, 1.2, dpi = 96, bg = "grey15") |>
  draw(text_grob("easy to read", gp = vl_gpar(fontsize = 22, col = "#DDDDDD"))))
```

**`cvd_collision` catches what the person reviewing the figure cannot.** Two
colours meant to be told apart, that a colour-blind reader sees as one. The
colours go through the same simulation matrices `render(cvd = )` draws with, and
distances are compared in Oklab:

```{r}
swatches <- function(cols) {
  s <- vl_scene(4, 1, dpi = 96, bg = "white")
  xs <- seq(0.1, 0.9, length.out = length(cols))
  for (i in seq_along(cols)) {
    s <- draw(s, rect_grob(x = xs[i], width = 0.12, height = 0.6,
                           gp = vl_gpar(fill = cols[i], col = NA)))
  }
  s
}

# ggplot2's default three-colour palette.
vl_lint(swatches(c("#F8766D", "#00BA38", "#619CFF")), rules = "cvd_collision")

# Okabe-Ito, designed to survive colour-vision deficiency.
vl_lint(swatches(c("#E69F00", "#56B4E9", "#009E73", "#CC79A7")),
        rules = "cvd_collision")
```

The threshold is calibrated so that the second call stays quiet: a rule that
flags a CVD-safe palette is a rule you would switch off. Point it at other
deficiencies with `cvd =`, where `"achromatopsia"` doubles as a greyscale-printer
check, or switch it off with `cvd = "none"`. See
`vignette("accessible-output")` for simulating the whole image.

**`font_fallback` reports characters no font on *this machine* can draw.** They
shape to glyph 0 and render as tofu boxes, and nothing about the label string
says so. The answer is deliberately machine-dependent, which makes it the rule
most worth running in CI — alongside `font_pin()` for the related question of
whether the fonts moved.

**`overplotted`** is the rule form of the `scene_stats()` measurement below,
measured per layer so it names the one to fix.

### Adding your own rules

The rule set is a registry, which is the point: vellum knows about geometry, but
it knows nothing about *your* semantics. A grammar layer can add rules that do —
a scale with one level, a legend with forty entries — and both come back from one
`vl_lint()` call.

A rule is a function of `(scene, nodes, ctx)` returning `vl_lint_finding()`, or
`NULL`. `nodes` is the resolved per-node table, in paint order:

```{r}
vl_lint_rule("show_columns", function(scene, nodes, ctx) {
  print(names(nodes))
  NULL
}, "prints its own input")

invisible(vl_lint(bad, rules = "show_columns"))
```

Boxes and clip boxes in device pixels, the viewport extent, the resolved fill and
stroke, stroke width, text size, and a `notdef` count. `ctx` carries the page
size, the dpi, the thresholds, and three lazy accessors: `pixel(x, y)`,
`region(x0, y0, x1, y1)` for a whole block of composited pixels, and `elements()`
for per-element geometry — which is what you want whenever a node is a batch,
since a scatter is *one* node whose box is the union over every point.

```{r}
vl_lint_rule(
  "wide_marks",
  function(scene, nodes, ctx) {
    wide <- nodes[(nodes$x1 - nodes$x0) > 0.8 * ctx$w, , drop = FALSE]
    vl_lint_finding("wide_marks", "note", wide, "spans almost the whole page")
  },
  description = "A mark spans nearly the full page width."
)

vl_lint(bad, rules = c("wide_marks", "tiny_text"))
```

Declaring `kinds` lets `vl_lint()` skip your rule entirely on a scene that
contains none of them, and `needs_pixels` tells a caller which rules cost a
render. A rule that fails does not take the lint down with it — it comes back as
a `rule_error` finding naming the rule, and everything else still runs.

Register from your package's `.onLoad()` and your users get your rules for free.

## `scene_stats()`: how crowded is this, really?

A mark count tells you nothing about whether the marks land on top of each
other. `scene_stats()` measures what actually reached the canvas:

```{r}
cloud <- function(n, sd = NA) {
  xy <- if (is.na(sd)) list(runif(n), runif(n)) else list(rnorm(n, 0.5, sd), rnorm(n, 0.5, sd))
  vl_scene(4, 3, dpi = 96) |>
    draw(points_grob(xy[[1]], xy[[2]], gp = vl_gpar(fill = "steelblue", col = NA)))
}

round(rbind(
  scattered = scene_stats(cloud(2000)),
  clustered = scene_stats(cloud(2000, sd = 0.03))
), 3)
```

Identical element counts; wildly different plots. `overplot` estimates how many
times an average inked pixel was painted over — `1` means no overlap — and it is
the honest trigger for reaching for `datashade()`. Thirty thousand well-separated
points are fine; eight thousand piled up are not, and only this number knows the
difference.

It is computed from bounding boxes, so treat it as an index rather than a
measurement: a circle fills about 79% of its box, and text and unkeyed
lines/polygons contribute to `ink` but not to `overplot`.

## `profile_render()`: where the time goes

```{r}
heavy <- vl_scene(6, 4, dpi = 96) |>
  draw(points_grob(runif(20000), runif(20000), name = "points")) |>
  draw(segments_grob(runif(2000), runif(2000), runif(2000), runif(2000), name = "edges")) |>
  draw(text_grob(format(1:200), x = runif(200), y = runif(200),
                 gp = vl_gpar(fontsize = 8), name = "labels"))

profile_render(heavy, reps = 1)
```

**Read the phase split first.** A render is three phases: *build* (constructing
the R value), *compile* (the R→Rust replay, including text shaping), and *raster*
(drawing). If compile dominates — which it does on scenes made of many small
grobs, where the cost is S7 validation and grob construction — then the backend
is not your problem and tuning it will not help. `profile_render()` says so
explicitly when that happens.

The per-mark table covers raster time, aggregated by mark rather than by node: a
vectorised `text_grob()` of 200 labels compiles to 200 nodes, and 200 rows of
almost nothing is noise rather than a profile.

Timing is armed only for that call, so ordinary renders pay nothing for it.

## Putting it in a test

All three return plain data frames, so they drop straight into a test suite.
`vl_lint_assert()` is the gate: it fails on warnings, reports what it found, and
passes everything else through to `vl_lint()`.

```{r}
#| eval: false
test_that("the figure has no lint findings", {
  vl_lint_assert(my_plot())
})
```

Real figures usually contain one deliberate oddity, and a linter you cannot
satisfy is a linter you turn off. `exclude` names the nodes you have already
looked at and accepted:

```{r}
#| eval: false
vl_lint_assert(my_plot(), exclude = "bleed_marker", severity = "note")
```

Suppression is by node rather than by rule, so the rule keeps working everywhere
else in the figure. An `exclude` entry that matches nothing warns — a stale
suppression list otherwise looks exactly like a working one.

Because `font_fallback` and `low_contrast` read the machine and the rendered
image, the CI run genuinely checks something the author's machine could not.

## Where to go next

- `vignette("render-quality")`: colour-vision simulation, crisp gridlines, and
  group effects — the other half of "will this actually read".
- `vignette("retained-mode")`: `scene_model()`, `hit_test()`, and editing nodes
  by name.
