---
title: "Placement: repelling labels, finding room, outlining groups"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Placement: repelling labels, finding room, outlining groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.4,
  dpi = 96,
  dev = "png"
)
library(vellum)
set.seed(3)
```

Overlapping labels are the most common defect in a data graphic, and the hardest
to fix from above the engine — because fixing them needs to know where
everything *ended up*, which is information that normally only exists after
drawing.

vellum resolves layout before it draws, so it knows. This article covers the
three things that follow from that: moving labels apart, finding somewhere to
put one thing, and outlining a group.

## Repelling labels

```{r}
n <- 26
x <- runif(n) * 0.9 + 0.05
y <- runif(n) * 0.85 + 0.08

scatter <- function() {
  vl_scene(6, 3.4, dpi = 96, bg = "white") |>
    draw(points_grob(x, y, size = vl_unit(2, "mm"),
                     gp = vl_gpar(fill = "#C0392B", col = NA), name = "pts")) |>
    draw(text_grob(paste0("station ", seq_len(n)), x = x, y = y,
                   gp = vl_gpar(fontsize = 8), name = "lab"))
}
display(scatter())
```

```{r}
display(vl_repel(scatter(), padding = 0.6))
```

Every label is legible, none has left its point, and nothing was dropped.

### It does not care what coordinate system you used

This is the part worth understanding, because it is what makes the feature
different from doing the same thing a layer up.

The solve happens in **device pixels**, over boxes that have already resolved.
The answer is then applied as an *absolute millimetre offset* on top of each
label's existing coordinate — vellum's compound `native + mm` unit. So the label
keeps its data anchor and moves by exactly the distance the solver asked for,
whatever produced that anchor.

A label in `native` units inside a panel with a scale of `0..100` on one axis and
`0..0.01` on the other moves by the same mechanism as one in `npc` on the page.
Panels are not solved one at a time; they are all just boxes.

```{r}
#| fig.height: 3
two <- vl_scene(6, 3, dpi = 96, bg = "white")
for (i in 1:2) {
  px <- runif(14) * 100
  py <- runif(14) * 0.01
  two <- two |>
    push(vl_viewport(name = paste0("p", i), x = c(0.25, 0.75)[i], width = 0.46,
                     xscale = c(0, 100), yscale = c(0, 0.01))) |>
    draw(rect_grob(gp = vl_gpar(fill = "grey97", col = "grey85"))) |>
    draw(points_grob(vl_unit(px, "native"), vl_unit(py, "native"),
                     size = vl_unit(1.6, "mm"),
                     gp = vl_gpar(fill = "#2C6FA6", col = NA))) |>
    draw(text_grob(paste0(c("a", "b")[i], seq_len(14)),
                   x = vl_unit(px, "native"), y = vl_unit(py, "native"),
                   gp = vl_gpar(fontsize = 7), name = paste0("lab", i))) |>
    pop()
}
display(vl_repel(two, padding = 0.5))
```

### The solve on its own

`vl_repel()` is `vl_place()` plus applying the answer. `vl_place()` hands you the
answer instead:

```{r}
sol <- vl_place(scatter(), padding = 0.6)
head(sol[, c("name", "index", "dx", "dy", "resolved")])
```

`dx`/`dy` are millimetres, and `resolved` says whether that label ended up clear
of everything.

Two reasons to want this rather than the finished scene. The first is leader
lines — the shift is a plain offset, so drawing one is arithmetic:

```{r}
moved <- which(abs(sol$dx) + abs(sol$dy) > 1)
with_leaders <- scatter() |>
  draw(segments_grob(
    x0 = vl_unit(x[moved], "npc"), y0 = vl_unit(y[moved], "npc"),
    x1 = vl_unit(x[moved], "npc") + vl_unit(sol$dx[moved], "mm"),
    y1 = vl_unit(y[moved], "npc") + vl_unit(sol$dy[moved], "mm"),
    gp = vl_gpar(col = "grey70", lwd = 0.6)
  ))
display(vl_repel(with_leaders, labels = "lab", padding = 0.6))
```

The second is that `resolved = FALSE` is a decision point. On a genuinely
crowded page some labels cannot be placed, and the useful responses — drop this
one, abbreviate it, shrink the font, show it on hover — all depend on knowing
what the labels *mean*. vellum reports the failure and stops there rather than
guessing; that call belongs to the layer above.

### How it works, and where it gives up

Relaxation, then repair. Overlaps push labels apart along the axis of least
separation; labels that are clear are eased back toward their anchor. That alone
is a local method, and local methods get stuck — a label slides out of the
collision it is in and never considers that the far side of its anchor was free
the whole time. So a second pass takes every label still colliding and tries
candidate positions on a ring around its anchor, nearest first, taking the first
that is clear. On a crowded scatter that pass is the difference between most
overlaps remaining and almost none.

`max_shift` bounds how far a label may travel (10 mm by default), so labels stay
near what they label rather than solving the puzzle by scattering.

## Where is there room?

The other placement question. Not "how do I move these apart" but "where can
this one thing go" — which is what automatic legend and annotation placement
needs.

```{r}
#| fig.height: 3.2
cloud <- vl_scene(5, 3.2, dpi = 96, bg = "white") |>
  draw(points_grob(rbeta(220, 2, 5), runif(220), size = vl_unit(1.6, "mm"),
                   gp = vl_gpar(fill = "#7FB2E5", col = NA)))
gap <- vl_empty_region(cloud, grid = 260)
round(gap, 3)
```

That composes neatly with width-constrained text
(`vignette("typography")`): the region also comes back in millimetres, which is
exactly the absolute measure `text_grob(width=)` wants — so an annotation can be
fitted to the gap that was just found, rather than to a width guessed in
advance.

```{r}
#| fig.height: 3.2
mm <- vl_empty_region(cloud, grid = 260, unit = "mm")
display(
  cloud |>
    draw(rect_grob(x = mean(gap[c("x0", "x1")]), y = mean(gap[c("y0", "y1")]),
                   width = gap[["x1"]] - gap[["x0"]],
                   height = gap[["y1"]] - gap[["y0"]],
                   gp = vl_gpar(fill = "#FFFFFFCC", col = "grey60"))) |>
    draw(text_grob("placed in the emptiest rectangle, and fitted to it",
                   x = mean(gap[c("x0", "x1")]), y = mean(gap[c("y0", "y1")]),
                   align = "centre", fit = TRUE,
                   width = vl_unit(mm[["x1"]] - mm[["x0"]] - 4, "mm"),
                   height = vl_unit(mm[["y1"]] - mm[["y0"]] - 4, "mm"),
                   gp = vl_gpar(fontsize = 11, col = "grey30")))
)
```

Occupancy is rasterised onto a grid, so the answer is exact *on that grid*. That
is a deliberate approximation — the exact maximal empty rectangle over n boxes
is superquadratic, and nothing is placed to sub-pixel tolerance. Boxes round
outward, so the result is conservative: it will never claim space that is in
fact occupied. Raise `grid` to trade time for precision.

## Outlining a group

`vl_hull()` and `vl_buffer()` are here rather than with the other geometry
because exclusion zones are what label placement wants them for.

```{r}
#| fig.height: 3.2
grp <- data.frame(
  x = c(rnorm(40, 0.32, 0.07), rnorm(40, 0.7, 0.06)),
  y = c(rnorm(40, 0.6, 0.09), rnorm(40, 0.38, 0.07)),
  g = rep(1:2, each = 40)
)
s <- vl_scene(5, 3.2, dpi = 96, bg = "white")
for (g in 1:2) {
  p <- grp[grp$g == g, ]
  h <- vl_hull(p$x, p$y, concavity = 4)
  b <- vl_buffer(h$x, h$y, 0.03)
  s <- s |>
    draw(polygon_grob(b$x, b$y,
                      gp = vl_gpar(fill = c("#E8F0F9", "#FBEDE7")[g], col = NA))) |>
    draw(polygon_grob(h$x, h$y,
                      gp = vl_gpar(fill = NA, col = c("#2C6FA6", "#C0392B")[g],
                                   lwd = 1.2, lty = "dashed"))) |>
    draw(points_grob(p$x, p$y, size = vl_unit(1.5, "mm"),
                     gp = vl_gpar(fill = c("#2C6FA6", "#C0392B")[g], col = NA)))
}
display(s)
```

`concavity` runs the opposite way to what the name suggests: **larger is more
convex**. `Inf` (the default) is exactly the convex hull, `8` follows the points
loosely, `4` is a good tight outline, and below about `3` the boundary starts
threading between interior points and crossing itself.

```{r}
#| fig.height: 1.8
p <- data.frame(x = rnorm(40, 0.5, 0.1), y = rnorm(40, 0.5, 0.12))
row <- vl_scene(6, 1.8, dpi = 96, bg = "white")
for (i in seq_along(cs <- c(Inf, 8, 4, 2))) {
  h <- vl_hull(p$x, p$y, concavity = cs[i])
  row <- row |>
    push(vl_viewport(x = (i - 0.5) / 4, width = 0.25)) |>
    draw(polygon_grob(h$x, h$y, gp = vl_gpar(fill = "#DCE7F5", col = "steelblue"))) |>
    draw(points_grob(p$x, p$y, size = vl_unit(1.2, "mm"),
                     gp = vl_gpar(fill = "grey25", col = NA))) |>
    draw(text_grob(paste0("concavity = ", cs[i]), y = 0.06,
                   gp = vl_gpar(fontsize = 8, col = "grey40"))) |>
    pop()
}
display(row)
```

Self-intersection at low `concavity` is inherent to the method, not a defect: a
boundary that threads between interior points has to cross itself eventually.
The same is true of buffering a concave ring — repairing either is a boolean
union, which vellum does not have yet.

## Where to go next

- `vignette("typography")`: the width-constrained text this pairs with.
- `vignette("inspecting-scenes")`: `vl_lint()`, which reports overlap and
  legibility problems rather than fixing them.
- `vignette("retained-mode")`: `edit_node()`, the mechanism `vl_repel()` uses to
  return a modified scene.
