---
title: "Render quality: accessibility, crisp edges, group effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Render quality: accessibility, crisp edges, group effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 2.2,
  dpi = 96,
  dev = "png"
)
library(vellum)
set.seed(1)
```

Three things that decide whether a figure reads correctly rather than merely
being correct: whether its colours survive a reader's vision, whether its thin
lines survive the pixel grid, and whether it can carry the depth cues designers
reach for.

## Colour-vision simulation

Roughly one man in twelve has some form of colour-vision deficiency, and the
red/green pair that most default palettes open with is exactly the one that
fails. `cvd =` re-renders a scene as such a reader would see it — on `render()`,
`scene_png()` and `scene_raster()`.

```{r}
pal <- c(red = "#D62728", green = "#2CA02C", blue = "#1F77B4",
         orange = "#FF7F0E", purple = "#9467BD")

swatches <- local({
  s <- vl_scene(6, 1.1, dpi = 96, bg = "white")
  n <- length(pal)
  for (i in seq_len(n)) {
    s <- draw(s, rect_grob(x = (i - 0.5) / n, width = 0.9 / n, height = 0.6, y = 0.6,
                           gp = vl_gpar(fill = pal[[i]], col = NA)))
    s <- draw(s, text_grob(names(pal)[i], x = (i - 0.5) / n, y = 0.12,
                           gp = vl_gpar(fontsize = 9, col = "grey30")))
  }
  s
})

# `scene_raster()` gives pixels back, so a simulation can be shown inline.
show_cvd <- function(scene, kind) {
  a <- scene_raster(scene, cvd = kind)
  m <- matrix(grDevices::rgb(a[1, , ], a[2, , ], a[3, , ], maxColorValue = 255),
              nrow = dim(a)[2])
  grid::grid.newpage()
  grid::grid.raster(as.raster(t(m)), interpolate = FALSE)
}
```

Normal vision:

```{r}
#| fig.height: 1.1
show_cvd(swatches, "none")
```

Deuteranopia, the common form:

```{r}
#| fig.height: 1.1
show_cvd(swatches, "deuteranopia")
```

Red and green have become the same colour. If those two encoded different
series, that plot no longer works.

The useful part is not the picture but the number. Because `scene_raster()`
returns pixels, you can *measure* separability rather than eyeball it:

```{r}
sep <- function(a, b, kind) {
  r <- scene_raster(swatches, cvd = kind)
  at <- function(i) as.integer(r[1:3, round((i - 0.5) / length(pal) * dim(r)[2]), 60])
  sum(abs(at(a) - at(b)))
}
vapply(c("none", "protanopia", "deuteranopia", "tritanopia", "achromatopsia"),
       function(k) sep(1, 2, k), 0)
```

`achromatopsia` doubles as a greyscale-printing check.

**And when the simulation says your palette fails, the fix is texture.** Encoding
a category by hatch angle as well as hue survives any of these simulations, and
greyscale printing too — see `vl_hatch()` in `vignette("scene-and-paint")`.

The simulation uses the Machado et al. (2009) matrices applied in **linear
light** — the common shortcut of applying them in sRGB shifts lightness as well
as hue. It is a raster post-pass: vector formats have no pixels to transform, and
`render()` warns rather than quietly writing an unsimulated file.

## Crisp gridlines

A one-pixel horizontal rule whose centre lands at a fractional coordinate covers
half of each of two pixel rows, and antialiasing renders it as two grey rows
instead of one black one. Multiply by every gridline and this is why plots often
look slightly muddy on screen.

`crisp = TRUE` snaps axis-parallel strokes onto the pixel grid:

```{r}
rule <- function(...) {
  vl_scene(4, 0.5, dpi = 96, bg = "white") |>
    draw(segments_grob(0.05, 0.5013, 0.95, 0.5013,
                       gp = vl_gpar(col = "black", lwd = 1, ...)))
}
# The darkest pixel in the column tells you whether the line is solid.
c(default = min(scene_raster(rule())[1, 100, ]),
  crisp = min(scene_raster(rule(crisp = TRUE))[1, 100, ]))
```

A non-zero minimum means the darkest pixel in that column is still grey: the
stroke is spread across two rows. Zero means it landed on one row, solid. (The
exact default value depends on where the fractional coordinate falls, which is
the problem.)

Diagonals are left alone — there is no grid to snap them to — and it only affects
raster output, since a vector format has no pixel grid.

Set it once on a viewport and every rule inside inherits it:

```{r}
#| eval: false
push(vl_viewport(gp = vl_gpar(crisp = TRUE)))
```

The companion is `antialias = FALSE`, for the cases where soft edges are wrong
rather than right: pixel art, QR codes, and heatmap cells that must tile without
a seam.

```{r}
tri <- function(aa) {
  vl_scene(1, 1, dpi = 96, bg = "white") |>
    draw(polygon_grob(c(0.1, 0.9, 0.5), c(0.1, 0.1, 0.9),
                      gp = vl_gpar(fill = "black", col = NA, antialias = aa)))
}
c(antialiased = length(unique(as.vector(scene_raster(tri(TRUE))[1, , ]))),
  aliased = length(unique(as.vector(scene_raster(tri(FALSE))[1, , ]))))
```

Two distinct values means pure black and pure white — no intermediate shades.

## Group effects

`blur` and `shadow` are **group** effects: they act on a viewport's contents
composited as one layer. That is the important distinction. Overlapping shapes
inside the viewport cast a single shadow together, rather than each casting one
onto the others.

```{r}
s <- vl_scene(6, 2, dpi = 96, bg = "grey97")
for (spec in list(
  list(x = 0.14, args = list()),
  list(x = 0.38, args = list(shadow = vl_shadow(dx = 2, dy = 3, blur = 4))),
  list(x = 0.62, args = list(blur = 2.5)),
  list(x = 0.86, args = list(shadow = vl_shadow(dx = 0, dy = 0, blur = 8, col = "#1F77B4")))
)) {
  s <- push(s, do.call(vl_viewport, c(list(x = spec$x, width = 0.2, height = 0.62), spec$args)))
  s <- draw(s, roundrect_grob(r = 0.14, gp = vl_gpar(fill = "steelblue", col = NA)))
  s <- pop(s)
}
display(s)
```

Left to right: plain, drop shadow, blur, and a glow — which is simply a shadow
with no offset, so it needs no separate function.

Per backend: the raster path convolves (three box passes, the standard Gaussian
approximation); SVG emits native `feGaussianBlur` / `feDropShadow`, so the group
stays vector and the viewer does the work; PDF has no filter model at all, and
says so through vellum's degradation warning rather than silently rasterising
your vector output.

```{r}
#| eval: false
render(s, "cards.pdf")
#> Warning: a group blur/shadow (PDF has no filter model; the group is drawn unfiltered)
```

That warning is the design principle at work: where a backend cannot honour
something, it fails visibly.

## Where to go next

- `vignette("inspecting-scenes")`: `vl_lint()` will flag low-contrast text and
  illegible labels for you, which pairs directly with the simulation above.
- `vignette("typography")`: halos and OpenType features.
