---
title: "Retained-mode superpowers: hit-testing and editable nodes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retained-mode superpowers: hit-testing and editable nodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
```

grid, and most drawing APIs, are *immediate mode*: a call paints pixels and
forgets everything about them. vellum is *retained mode*. The scene you build
is kept as a tree until you render it, and that tree can be queried and edited
afterwards. It buys you three things: naming and
editing nodes, hit-testing, and reading back a per-element model of the scene.

## Naming nodes

Every grob and viewport takes an optional `name`. A name turns a node into
something you can look up and modify later.

```{r}
#| label: build
dots <- vl_scene(5, 3, bg = "white") |>
  draw(circle_grob(x = 0.25, y = 0.5, r = 0.16, name = "a",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA))) |>
  draw(circle_grob(x = 0.55, y = 0.5, r = 0.16, name = "b",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA))) |>
  draw(circle_grob(x = 0.85, y = 0.5, r = 0.16, name = "c",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA)))

node_names(dots)
```

`node_names()` lists the names in paint order, and `get_node()` returns the node
itself, so you can inspect a value you built earlier.

```{r}
#| label: get
get_node(dots, "b")
```

## Editing a node

`edit_node()` returns a *new* scene with one node's properties changed. It is
copy-on-modify: the original scene value is untouched, so you can derive
variants without disturbing the source. Here we highlight the middle dot.

```{r}
#| label: edit
highlighted <- edit_node(dots, "b", gp = vl_gpar(fill = "#e74c3c", col = NA))
highlighted
```

The original is unchanged:

```{r}
#| label: original
dots
```

This is the mechanism a host uses for hover and selection: keep one scene, and
on an interaction re-derive it with the touched node restyled, then re-render.
Flagging that node's viewport with `cache = TRUE` (a repaint boundary) makes the
re-render cheap, since only the changed subtree is re-rasterised.

## Hit-testing

`hit_test()` answers the inverse question: given a point, which node is drawn on
top there? grid offers only `grid.locator()`, but a retained scene can be
compiled into a colour pick-buffer, so the answer is exact with respect to
geometry, clipping, and paint order. Coordinates default to `"npc"` (`0..1`,
y up); pass `units = "px"` for device pixels.

```{r}
#| label: hit
hit_test(dots, x = 0.25, y = 0.5) # over dot "a"
hit_test(dots, x = 0.55, y = 0.5) # over dot "b"
hit_test(dots, x = 0.05, y = 0.1) # empty space
```

A point over a named grob returns its name; over an unnamed grob it returns
`NA`; over blank canvas it returns `NULL`. That is enough to route a click back
to the datum that drew the mark.

## A per-element model of the scene

`hit_test()` picks one node; `scene_model()` returns the whole picture. It walks
a rendered scene and returns one row per drawn element of the *keyable* marks
(points, circles, rects, hexagons, sectors, segments), pairing each element's
identity with its resolved device-pixel bounding box.

```{r}
#| label: model
sm <- scene_model(dots)
str(sm, max.level = 1)
sm$elements[, c("mark", "name", "x", "y", "w", "h")]
```

The real power shows up when marks carry a data `key` (and optional free-form
`meta`), which the batched grobs accept per element. The `key` is emitted by the
SVG backend as `data-key` on each element and surfaced here, so a host can render
the SVG once (`scene_svg()`), then use this table to map a DOM event back to the
originating row of data.

```{r}
#| label: keyed
keyed <- vl_scene(5, 3, bg = "white") |>
  draw(points_grob(
    x = vl_unit(c(0.25, 0.55, 0.85), "npc"), y = 0.5,
    size = vl_unit(6, "mm"),
    key = c("row-1", "row-2", "row-3"),
    gp = vl_gpar(fill = "#3a7bd5", col = NA)
  ))

scene_model(keyed)$elements[, c("mark", "key", "x", "y")]
```

## Why this matters

The retained scene graph is what separates vellum from an immediate-mode
drawing layer. Because the tree survives past construction:

- nodes can be **named, inspected, and edited** (`node_names()`, `get_node()`,
  `edit_node()`) without rebuilding the scene;
- any point can be **hit-tested** back to the node that drew it (`hit_test()`);
- the whole scene can be read back as a **table of elements with keys and
  geometry** (`scene_model()`), the host-agnostic bridge to interactivity.

These are the primitives an interactive grammar layer builds tooltips, brushing,
and linked selection on top of.
```
