---
title: "Retained-mode scenes: editing, hit-testing, and read-back"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retained-mode scenes: editing, hit-testing, and read-back}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
```

vellum is *retained mode*: the scene you build is kept as a tree until you
render it, and that tree can be queried and edited afterwards. It gives you
three things: naming and editing nodes, hit-testing, and reading back a
per-element model of the scene.

grid is retained too, so the first of those three is not new. A `gTree` is a
value, `editGrob()` derives a modified copy of one, and a drawn plot's display
list can be walked and rewritten with `grid.ls()`, `grid.get()`, and
`grid.edit()`. What differs here is that the tree is the primary artefact rather
than device state you recover, and that the scene can answer questions about
where things landed: which node is under this point, and what bounding box did
each element get. Those two grid leaves to you.

## Naming nodes

Every grob and viewport takes an optional `name`. A name turns a node into
something you can look up and modify later.

```{r}
#| label: build
dots <- vl_scene(5, 3, bg = "white") |>
  draw(circle_grob(x = 0.25, y = 0.5, r = 0.16, name = "a",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA))) |>
  draw(circle_grob(x = 0.55, y = 0.5, r = 0.16, name = "b",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA))) |>
  draw(circle_grob(x = 0.85, y = 0.5, r = 0.16, name = "c",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA)))

node_names(dots)
```

`node_names()` lists the names in paint order, and `get_node()` returns the node
itself, so you can inspect a value you built earlier.

```{r}
#| label: get
get_node(dots, "b")
```

## Editing a node

`edit_node()` returns a *new* scene with one node's properties changed. It is
copy-on-modify: the original scene value is untouched, so you can derive
variants without disturbing the source. `editGrob()` gives you the same guarantee
in grid; what `edit_node()` adds is that untouched subtrees keep their internal
node ids, which is what the repaint-boundary cache keys on. Both walk the tree to
find the node, but neither copies a subtree it did not touch: on a scene of
20,000 named sibling grobs an edit costs about 14 ms here against about 430 ms for
`editGrob()`, because grid's copy duplicates the siblings and this one does not.
Here we highlight the middle dot.

```{r}
#| label: edit
highlighted <- edit_node(dots, "b", gp = vl_gpar(fill = "#e74c3c", col = NA))
highlighted
```

The original is unchanged:

```{r}
#| label: original
dots
```

This is the mechanism a host uses for hover and selection: keep one scene, and
on an interaction re-derive it with the touched node restyled, then re-render.
Flagging that node's viewport with `cache = TRUE` (a repaint boundary) makes the
re-render cheap, since only the changed subtree is re-rasterised.

## Hit-testing

`hit_test()` answers the inverse question: given a point, which node is drawn on
top there? grid ships `grid.locator()`, which waits for one interactive click and
returns coordinates rather than a node; you can build a picker on top of
`grobPoints()` and a point-in-polygon test, or tag exported SVG elements the way
gridSVG and ggiraph do. vellum compiles the retained scene into a colour
pick-buffer instead, through the same transform, clipping, and paint-order code
that drew it, so the answer is exact with respect to geometry, clipping, and
paint order rather than coming from a second geometry implementation you have to
keep in sync. It is also an ordinary function call: no device, no user, callable
as often as you like. Coordinates default to `"npc"` (`0..1`, y up); pass
`units = "px"` for device pixels.

```{r}
#| label: hit
hit_test(dots, x = 0.25, y = 0.5) # over dot "a"
hit_test(dots, x = 0.55, y = 0.5) # over dot "b"
hit_test(dots, x = 0.05, y = 0.1) # empty space
```

A point over a named grob returns its name; over an unnamed grob it returns
`NA`; over blank canvas it returns `NULL`. That is enough to route a click back
to the datum that drew the mark.

## A per-element model of the scene

`hit_test()` picks one node; `scene_model()` returns the whole picture. It walks
a rendered scene and returns one row per drawn element of the *keyable* marks
(points, circles, rects, hexagons, sectors, segments), pairing each element's
identity with its resolved device-pixel bounding box.

```{r}
#| label: model
sm <- scene_model(dots)
str(sm, max.level = 1)
sm$elements[, c("mark", "name", "x", "y", "w", "h")]
```

The real power shows up when marks carry a data `key` (and optional free-form
`meta`), which the batched grobs accept per element. The `key` is emitted by the
SVG backend as `data-key` on each element and surfaced here, so a host can render
the SVG once (`scene_svg()`), then use this table to map a DOM event back to the
originating row of data.

```{r}
#| label: keyed
keyed <- vl_scene(5, 3, bg = "white") |>
  draw(points_grob(
    x = vl_unit(c(0.25, 0.55, 0.85), "npc"), y = 0.5,
    size = vl_unit(6, "mm"),
    key = c("row-1", "row-2", "row-3"),
    gp = vl_gpar(fill = "#3a7bd5", col = NA)
  ))

scene_model(keyed)$elements[, c("mark", "key", "x", "y")]
```

## Why this matters

Because the tree survives past construction as a plain value:

- nodes can be **named, inspected, and edited** (`node_names()`, `get_node()`,
  `edit_node()`) without rebuilding the scene, as they can in grid;
- any point can be **hit-tested** back to the node that drew it (`hit_test()`);
- the whole scene can be read back as a **table of elements with keys and
  geometry** (`scene_model()`), the host-agnostic bridge to interactivity.

These are the primitives an interactive grammar layer builds tooltips, brushing,
and linked selection on top of.
```
