---
title: "Asking a scene what it drew: read-back, picking, and editing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Asking a scene what it drew: read-back, picking, and editing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
```

A vellum scene is a value, and once it has been through the layout solve it can
answer questions about itself: where did each element end up, and what is drawn
at this point? That read-back is the capability with no counterpart in grid, and
it is what an interactive host, an accessibility layer, or a test suite binds to.
This vignette covers it, and the editing API that goes with it.

Being retained is not the interesting part. grid retains its scene too: a `gTree`
is a value, `editGrob()` derives a modified copy, and a drawn plot's display list
can be walked and rewritten with `grid.ls()`, `grid.get()`, and `grid.edit()`.
Naming and editing nodes below will look familiar for that reason. What grid
leaves to you is everything in the first paragraph, because its resolved geometry
exists only during a draw and is gone when the draw ends.

## Naming nodes

Every grob and viewport takes an optional `name`. A name turns a node into
something you can look up and modify later.

```{r}
#| label: build
dots <- vl_scene(5, 3, bg = "white") |>
  draw(circle_grob(x = 0.25, y = 0.5, r = 0.16, name = "a",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA))) |>
  draw(circle_grob(x = 0.55, y = 0.5, r = 0.16, name = "b",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA))) |>
  draw(circle_grob(x = 0.85, y = 0.5, r = 0.16, name = "c",
                   gp = vl_gpar(fill = "#bdc3c7", col = NA)))

node_names(dots)
```

`node_names()` lists the names in paint order, and `get_node()` returns the node
itself, so you can inspect a value you built earlier.

```{r}
#| label: get
get_node(dots, "b")
```

## Editing a node

`edit_node()` returns a *new* scene with one node's properties changed. It is
copy-on-modify: the original scene value is untouched, so you can derive
variants without disturbing the source. `editGrob()` gives you the same guarantee
in grid; what `edit_node()` adds is that untouched subtrees keep their internal
node ids, which is what the repaint-boundary cache keys on. Both walk the tree to
find the node, but neither copies a subtree it did not touch: on a scene of
20,000 named sibling grobs an edit costs about 14 ms here against about 430 ms for
`editGrob()`, because grid's copy duplicates the siblings and this one does not.
Here we highlight the middle dot.

```{r}
#| label: edit
highlighted <- edit_node(dots, "b", gp = vl_gpar(fill = "#e74c3c", col = NA))
highlighted
```

The original is unchanged:

```{r}
#| label: original
dots
```

This is the mechanism a host uses for hover and selection: keep one scene, and
on an interaction re-derive it with the touched node restyled, then re-render.
Flagging that node's viewport with `cache = TRUE` (a repaint boundary) makes the
re-render cheap, since only the changed subtree is re-rasterised.

## Hit-testing

`hit_test()` answers the inverse question: given a point, which node is drawn on
top there? grid ships `grid.locator()`, which waits for one interactive click and
returns coordinates rather than a node; you can build a picker on top of
`grobPoints()` and a point-in-polygon test, or tag exported SVG elements the way
gridSVG and ggiraph do. vellum compiles the retained scene into a colour
pick-buffer instead, through the same transform, clipping, and paint-order code
that drew it, so the answer is exact with respect to geometry, clipping, and
paint order rather than coming from a second geometry implementation you have to
keep in sync. It is also an ordinary function call: no device, no user, callable
as often as you like. Coordinates default to `"npc"` (`0..1`, y up); pass
`units = "px"` for device pixels.

```{r}
#| label: hit
hit_test(dots, x = 0.25, y = 0.5) # over dot "a"
hit_test(dots, x = 0.55, y = 0.5) # over dot "b"
hit_test(dots, x = 0.05, y = 0.1) # empty space
```

A point over a named grob returns its name; over an unnamed grob it returns
`NA`; over blank canvas it returns `NULL`. That is enough to route a click back
to the datum that drew the mark.

## A per-element model of the scene

`hit_test()` picks one node; `scene_model()` returns the whole picture. It walks
a rendered scene and returns one row per drawn element, pairing each element's
identity with its resolved device-pixel bounding box.

```{r}
#| label: model
sm <- scene_model(dots)
str(sm, max.level = 1)
sm$elements[, c("mark", "name", "x", "y", "w", "h")]
```

The real power shows up when marks carry a data `key` (and optional free-form
`meta`), which the batched grobs accept per element. The `key` is emitted by the
SVG backend as `data-key` on each element and surfaced here, so a host can render
the SVG once (`scene_svg()`), then use this table to map a DOM event back to the
originating row of data.

```{r}
#| label: keyed
keyed <- vl_scene(5, 3, bg = "white") |>
  draw(points_grob(
    x = vl_unit(c(0.25, 0.55, 0.85), "npc"), y = 0.5,
    size = vl_unit(6, "mm"),
    key = c("row-1", "row-2", "row-3"),
    gp = vl_gpar(fill = "#3a7bd5", col = NA)
  ))

scene_model(keyed)$elements[, c("mark", "key", "x", "y")]
```

### Every mark family can be keyed

Points, rects, circles, hexagons, sectors and segments are *batched* marks: they
report a row whether or not they carry a key, so a plain scene still yields a
geometry table.

Lines, polygons, paths, rounded rects and text carry keys too, and report a row
**only when keyed**. The asymmetry is deliberate: a plot is full of unkeyed
gridlines and axis labels, and reporting those would bury the marks that mean
something in thousands of phantom elements.

```{r}
#| label: keyed-families
series <- vl_scene(5, 2.4, bg = "white") |>
  push(vl_viewport(name = "panel", xscale = c(1, 6), yscale = c(0, 10))) |>
  draw(lines_grob(vl_unit(1:6, "native"), vl_unit(c(2, 5, 4, 8, 6, 9), "native"),
                  gp = vl_gpar(col = "#2C6FA6", lwd = 2),
                  key = "series-A", meta = list(list(series = "A")))) |>
  draw(text_grob(c("low", "high"), x = vl_unit(c(1, 6), "native"),
                 y = vl_unit(c(2, 9), "native"), gp = vl_gpar(fontsize = 8),
                 key = c("lbl-low", "lbl-high"))) |>
  pop()

scene_model(series)$elements[, c("mark", "key", "x0", "x1")]
```

A whole series as one addressable thing — hover the line, highlight the series —
and per-label keys on the annotations. Neither was possible before: the marks
existed, but nothing could refer to them.

## Hit-testing that respects the shape

A bounding box is the right answer for a rectangular brush and the wrong one for
anything diagonal or thin. The line above runs from the bottom-left of its panel
to the top-right, so **its bounding box is almost the whole panel**:

```{r}
#| label: bbox-problem
el <- scene_model(series)$elements
el[el$key == "series-A", c("x0", "y0", "x1", "y1")]
```

Ask "what is nearest?" with only that, and the line answers from anywhere in the
plot. `vl_nearest()` measures to the geometry instead:

```{r}
#| label: nearest
demo <- vl_scene(4, 3, bg = "white") |>
  draw(segments_grob(0.12, 0.12, 0.88, 0.88,
                     gp = vl_gpar(col = "#C0392B", lwd = 2), key = "diagonal")) |>
  draw(points_grob(0.82, 0.18, size = vl_unit(3, "mm"),
                   gp = vl_gpar(fill = "#2C6FA6", col = NA), key = "corner"))

# This probe is INSIDE the diagonal's bounding box, and far from the diagonal.
vl_nearest(demo, 0.82, 0.18, n = 2)
```

Distances are to the real shape: perpendicular to a segment, to the disc for
round marks, to the nearest edge of an open polyline, and **zero anywhere inside**
a closed polygon or path — so clicking the middle of a choropleth region hits the
region rather than its nearest border.

```{r}
#| label: inside
region <- vl_scene(3, 3, bg = "white") |>
  draw(polygon_grob(c(0.2, 0.8, 0.5), c(0.2, 0.2, 0.8),
                    gp = vl_gpar(fill = "#F1C40F", col = "grey30"), key = "tri"))
vl_nearest(region, 0.5, 0.4)
```

### For a host that cannot ask

A browser cannot call back into R on every mouse move. `element_geometry()` hands
over the same true geometry **once**, so the client computes distances locally at
whatever rate it likes — typically an R-tree over the boxes to shortlist
candidates, then an exact test against these vertices to rank them.

```{r}
#| label: geometry
element_geometry(demo)
```

Two endpoints for the segment and one centre for the point — not four box
corners. That is the difference that lets a client hit-test a diagonal at all.

## Why this matters

Ranked by how much of it you cannot get elsewhere:

- the scene can be read back as a **table of elements with keys and geometry**
  (`scene_model()`), which is the host-agnostic bridge an interactive layer,
  a screen-reader description, or a positional test binds to;
- **any mark can be made addressable** — including lines, polygons and labels,
  which is what lets a whole series or an annotation be a thing the user can
  point at;
- marks can be picked by their **true geometry** (`vl_nearest()`), or that
  geometry handed to a client to do it itself (`element_geometry()`), so a
  diagonal is not matched from the far corner of its bounding box;
- **any point can be picked** back to the node that drew it (`hit_test()`),
  through the same compile path that rendered the scene, so the answer agrees
  with the picture by construction;
- nodes can be **named, inspected, and edited** (`node_names()`, `get_node()`,
  `edit_node()`) — as they can in grid, the difference being that you are editing
  a value rather than device state.

The first two are what `vellumwidget` is built on. It adds no drawing code of its
own: it renders the scene's SVG, reads this table, and attaches behaviour.
```
