---
title: "Scenes as values: saving, comparing, composing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scenes as values: saving, comparing, composing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
set.seed(1)
```

A vellum scene is an immutable value describing what to draw. This article is
about treating it as one: writing it to a file, fingerprinting it, comparing two
of them, and nesting one inside another.

None of this needs a device, because none of it needs pixels.

```{r}
plot_of <- function(fill = "steelblue", r = 0.3) {
  set.seed(1)
  vl_scene(4, 3, dpi = 96, bg = "white") |>
    push(vl_viewport(name = "panel", xscale = c(0, 1), yscale = c(0, 1))) |>
    draw(rect_grob(gp = vl_gpar(fill = "grey96", col = "grey70"))) |>
    draw(points_grob(runif(60), runif(60), gp = vl_gpar(fill = fill, col = NA))) |>
    draw(circle_grob(r = r, name = "marker",
                     gp = vl_gpar(fill = NA, col = "grey20", lwd = 2))) |>
    pop()
}
```

## Saving a scene

`scene_write()` and `scene_read()` persist a built scene and rebuild it. `.rds`
needs nothing; `.json` is a portable text format and needs jsonlite.

```{r}
f <- tempfile(fileext = ".rds")
scene_write(plot_of(), f)
identical(as_scene_spec(scene_read(f)), as_scene_spec(plot_of()))
```

What this is for: caching a built scene without re-running the code that built
it, sending one to a worker process to render, and keeping a reproducible
intermediate between the code and the pixels.

Underneath is `as_scene_spec()`, which turns the scene into a plain nested list
of R vectors, and `from_scene_spec()`, which rebuilds it. The conversion is
generic over the S7 property model rather than written per grob type, so a grob
added to vellum later serializes with no change to the serializer.

This sits one level below `vellumplot`'s *plot* spec, and the two compose rather
than compete. A plot spec is portable and re-renderable at any size; a scene spec
is the resolved artifact — the grobs, units and viewports as they will actually
be drawn. Keep the plot spec to re-render, keep the scene spec to reproduce
exactly this scene.

One deliberate omission: the build-time `nid` token is not serialized. It is a
monotonic counter used by the render cache, and keeping it would make two
identical scenes built separately serialize — and therefore hash and diff —
differently, which is the opposite of what a fingerprint is for.

## Fingerprinting

```{r}
c(
  same_content = scene_hash(plot_of()) == scene_hash(plot_of()),
  changed_fill = scene_hash(plot_of()) == scene_hash(plot_of(fill = "tomato"))
)
```

Two independently-built scenes with the same content hash the same; any change
to what is drawn changes the hash. That makes it usable as a cache key, and as a
test assertion that a refactor did not alter a figure.

## Comparing

`scene_hash()` answers *whether* something changed. `scene_diff()` answers
*what*:

```{r}
scene_diff(plot_of(), plot_of(fill = "tomato", r = 0.42))
```

This is a better basis for visual-regression testing than comparing rendered
images, for one specific reason: **an image diff is sensitive to the font
stack.** Run the same code on two machines with different fonts and the pixel
differences swamp the change you were looking for. A structural diff compares
what was *asked for*, and is unaffected.

```{r}
#| eval: false
test_that("the refactor did not change the figure", {
  expect_equal(nrow(scene_diff(reference_scene(), my_plot())), 0L)
})
```

## Composing

Because a scene has a known resolved size, nesting one inside another is a graft
rather than a re-render. `scene_inset()` splices the guest's grobs under a
viewport of the host:

```{r}
overview <- plot_of(fill = "grey65", r = 0.45)
detail <- vl_scene(1, 1) |>
  push(vl_viewport(xscale = c(0, 1), yscale = c(0, 1))) |>
  draw(rect_grob(gp = vl_gpar(fill = "white", col = "grey40"))) |>
  draw(points_grob(runif(30) * 0.4 + 0.3, runif(30) * 0.4 + 0.3,
                   gp = vl_gpar(fill = "tomato", col = NA))) |>
  pop()

display(
  scene_inset(overview, detail, x = 0.78, y = 0.76, width = 0.34, height = 0.4,
              name = "inset", shadow = vl_shadow(dx = 2, dy = 2, blur = 3))
)
```

The result is an ordinary scene. The inset is an addressable node, so it can be
found with `node_names()`, edited with `edit_node()`, inset into something else,
or serialized like anything else.

Two things to know. The guest's page size, background and dpi are *not* carried
over — it becomes a region of the host, which owns those. And the guest's `npc`
coordinates become relative to the region, so a guest built on a square page and
dropped into a wide region will stretch; match the aspect, or build the guest at
the aspect you want.

What vellum deliberately does *not* decide is the **policy**: whether panel edges
should align across composed plots, whether axes should be shared. Those are
properties of how a grammar decomposes a figure, so they belong in a layer above.
What the engine owes that layer is the ability to nest scenes at all.

## Where to go next

- `vignette("inspecting-scenes")`: `vl_lint()`, `scene_stats()` and
  `profile_render()` — the other things a resolved scene can tell you.
- `vignette("retained-mode")`: reading geometry back and editing nodes by name.
