---
title: "The scene contract: what quill and gloss depend on"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The scene contract: what quill and gloss depend on}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
library(vellum)
```

vellum is the backend for a small ecosystem: [quill](https://schochastics.github.io/quill/)
compiles a grammar of graphics *into* a vellum scene, and
[gloss](https://schochastics.github.io/gloss/) turns a rendered scene into an
interactive HTML widget. Both bind to a narrow seam — `scene_model()`,
`scene_svg()`, and the per-element metadata that rides on grobs. Because the
three packages are version-locked and co-released, a change to that seam can
break the layers above it silently.

**This article is the authoritative description of that seam.** If you are
building on vellum, or changing vellum's output, this is the contract to hold
to. (It supersedes the internal design note `_docs/DESIGN-INTERACTIVITY.md`,
which is kept only as history.)

## The seam, in one picture

```
  quill  ──compiles a plot to grobs carrying──▶  key / meta / id / role
                                                        │
  vellum  ──renders──▶  scene_svg()   (data-key, data-vellum-*, role attrs)
                        scene_model()  (per-element table: identity + bbox)
                                                        │
  gloss  ──reads scene_svg() + scene_model()──▶  hover / select / brush / link
```

Everything below the grammar is *pure metadata*: a static PNG/SVG render never
calls `scene_model()`, and a grob without a `key`/`meta` still draws exactly as
before. Nothing here changes what is painted.

## Carrying identity on a grob

The batched mark constructors — `points_grob()`, `circle_grob()`, `rect_grob()`,
`segments_grob()`, `hexagon_grob()`, `sector_grob()` — take two optional
per-element arguments:

- `key`: a character vector of **data keys**, one per element (recycled). This
  is the join key a host uses to tie an on-screen element back to a datum.
- `meta`: a list of **free-form per-element records** (recycled), e.g. a tooltip
  string or field values.

Every grob also takes `id` and `role` (a single value per grob) for semantic
identity and accessibility.

```{r}
#| label: build
s <- vl_scene(5, 3, bg = "white") |>
  draw(points_grob(
    x    = c(0.25, 0.5, 0.75),
    y    = c(0.4, 0.7, 0.4),
    size = unit(6, "mm"),
    gp   = gpar(fill = "#3498db", col = NA),
    key  = c("a", "b", "c"),
    meta = list(
      list(tooltip = "first"),
      list(tooltip = "second"),
      list(tooltip = "third")
    )
  ))
s
```

(Each batched grob carries one graphical style; the grammar layer above vellum
emits one grob per style bucket, so a scatter with three colours is three
`points_grob()`s. The `key`/`meta` machinery is per element regardless.)

## `scene_model()` — the per-element table

`scene_model()` walks the rendered scene and returns a list of **two data
frames**.

```{r}
#| label: model
m <- scene_model(s)
names(m)
```

### `elements`

One row per drawn element of the *keyable* marks, **in paint order**:

```{r}
#| label: elements
str(m$elements)
```

The columns are the contract:

| column | type | meaning |
|---|---|---|
| `key` | character | the data key (`NA` if the element was drawn without one) |
| `mark` | character | the mark kind (see vocabulary below) |
| `id` | character | the grob `id` (`NA` if unset) |
| `name` | character | the grob `name` (`NA` if unset) |
| `panel` | character | the enclosing named panel (`NA` if none) |
| `x0, y0, x1, y1` | numeric | device-pixel bounding box |
| `x, y` | numeric | element centre, `(x0+x1)/2, (y0+y1)/2` |
| `w, h` | numeric | element size, `x1-x0, y1-y0` |
| `meta` | list | the free-form per-element record (list-column) |

The `mark` vocabulary is a closed set:
`rect`, `point`, `circle`, `hexagon`, `sector`, `segment`, `path`, `line`,
`polygon`.

Two families produce these rows:

- **Batched marks** (`rect`, `point`, `circle`, `hexagon`, `sector`, `segment`)
  emit **one row per element, always** — even when unkeyed (then `key` is `NA`).
- **Single-shape marks** (`path`, `line`, `polygon`) emit **one row per grob, and
  only when keyed**. An unkeyed path/line/polygon is geometry-only and does not
  appear in the table. (A single `sf` feature — one polygon or linestring — is
  exactly one such element.) These constructors do not take a `key` argument; the
  grammar keys them by setting the grob's `keys` slot, which is how a single sf
  feature becomes addressable.

### `panels`

One row per **named panel** (a named viewport becomes an addressable panel), with
the bounding box of that panel's elements:

```{r}
#| label: panels
str(m$panels)
```

Columns: `name`, `x0`, `y0`, `x1`, `y1`.

### The `meta` key vocabulary

vellum treats `meta` as opaque: it is a free-form list, recycled to the element
count and carried through untouched — vellum does **not** name or validate its
keys. The following key names are **conventions** the grammar (quill) writes and
the widget (gloss) reads. Documenting them here keeps the three layers in step;
they are not enforced by vellum.

| `meta` key | written by | read by | purpose |
|---|---|---|---|
| `tooltip` | quill `tooltip=` | gloss | hover tooltip text (falls back to `key`) |
| `data_id` | quill `data_id=` | (becomes the `key`) | the data key for an element |
| `hover_group` | quill `hover_group=` | gloss | co-highlight a group on hover |
| `hover_color` | quill mark aesthetic | gloss | per-element hover outline colour |
| `selected_color` | quill mark aesthetic | gloss | per-element selection colour |
| `legend` | quill legend keying | gloss | the series a mark belongs to |
| `legend_for` | quill legend keying | gloss | the series a legend swatch drives (`"<aes>:<level>"`) |

If you add a convention, add it here.

## `scene_svg()` — the emitted attributes

`scene_svg()` returns the same scene as an SVG string. Interactivity rides on SVG
attributes; the raster and PDF backends ignore all of them.

```{r}
#| label: svg
svg <- scene_svg(s)
# the data-key attributes on the three points:
regmatches(svg, gregexpr('data-key="[^"]*"', svg))[[1]]
```

Four attribute mechanisms, each emitted only when its source is set:

| attribute | source | scope | set from |
|---|---|---|---|
| `data-key` | grob `key` | per element | `points_grob(key=)` etc. |
| `data-vellum-id` | grob `id` | per grob/node | `*_grob(id=)` |
| `data-vellum-name` | grob `name` | per grob/node | `*_grob(name=)` |
| `role` | grob `role` | per grob/node | `*_grob(role=)` |
| `data-vellum-panel` | named viewport | wrapping `<g>` | `viewport(name=)` |

An empty/`NULL` source emits **no attribute**, so a scene that declares no
interactivity is byte-for-byte identical to one built before any of this existed.

## Accessibility

A scene carries an optional accessible **name** and **description**, set with
`vl_scene(title=, desc=)` or `describe(scene, title=, desc=)`. When present:

- the **SVG** root becomes `<svg role="img" aria-labelledby="…">` with `<title>`
  and `<desc>` children (the reliable screen-reader pattern; WCAG 1.1.1);
- the **PDF** is a **tagged PDF** — the chart is a `Figure` in the structure tree
  whose `Alt` is the description.

This is additive: with no title/desc the output is unchanged. The grammar layer
(`quill`) sets these automatically from the plot's title and an auto-generated (or
`labs(alt=)`) alt text — see quill's [Accessibility
article](https://schochastics.github.io/quill/articles/accessibility.html).
Per-element `role` (above) is a genuine ARIA role; the interactive widget layer
(`gloss`) adds
keyboard navigation and live-region announcements on top.

## Invariants the contract guarantees

These are the properties quill and gloss are entitled to rely on, and that
vellum's own tests (`tests/testthat/test-contract.R`) pin down:

1. **Paint order.** `scene_model()$elements` is in draw order, and the SVG emits
   elements in the same order. A host can zip the SVG DOM and the table
   positionally.
2. **Semantic/geometry agreement.** `scene_model()` builds the identity columns
   from the R grob tree and the geometry columns from the compiled backend, then
   asserts they enumerate the same elements: the element counts must match and
   the `key` column must be identical at every position. A mismatch is a hard
   error (a compiler bug), never a silent mis-join.
3. **Additivity.** `key`/`meta`/`id`/`role` never change what is drawn. A render
   with no keys is identical to the same scene without the machinery, on every
   backend.
4. **Stable `id` join key.** A grob's `id` surfaces as `data-vellum-id` in the
   SVG *and* is the join key quill records in its provenance table, so a widget
   can map an SVG node to the grammar record that produced it.

## For contributors

If you change any of the following, update this article and the contract tests
in the same commit, and expect to co-release quill and gloss:

- the `elements`/`panels` column set or types,
- the `mark` vocabulary,
- which marks are keyable, or the batched-vs-single-shape rule,
- the SVG attribute names, or when they are emitted,
- the reserved `meta` key vocabulary.
```
