---
title: "The scene graph and the paint model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The scene graph and the paint model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 2.6,
  dpi = 96,
  dev = "png"
)
library(vellum)
```

This article covers the two pieces you spend most of your time with in vellum:
the **scene graph** (units, viewports, and the tree they form) and the **paint
model** (gradients, patterns, and masks) shared across every backend.

## The scene graph

A vellum scene is a tree. The root is the page created by `vl_scene()`; every
`push()` adds a `vl_viewport()` child and descends into it; every `draw()` appends
a grob at the current level; `pop()` climbs back up. The tree is retained rather
than drawn-and-forgotten, which is what enables the queries in
`vignette("retained-mode")`.

Because it is a tree, viewports nest, and a child's geometry is expressed
relative to its parent. That is the whole mechanism behind panels, insets, and
faceting: push a viewport for a sub-region, draw inside it in local
coordinates, then pop.

```{r}
#| label: nesting
#| fig.width: 6
#| fig.height: 2.4
vl_scene(6, 2.4, bg = "white") |>
  # a full-width band
  draw(rect_grob(height = 0.6, gp = vl_gpar(fill = "#eef2f6", col = NA))) |>
  # an inset viewport occupying the middle third
  push(vl_viewport(x = 0.5, width = 1 / 3, height = 0.8)) |>
  draw(rect_grob(gp = vl_gpar(fill = "#3a7bd5", col = NA))) |>
  draw(text_grob("inset", gp = vl_gpar(col = "white", fontface = "bold"))) |>
  pop()
```

## Units

Coordinates and sizes are `vl_unit()` vectors: a value paired with a unit name.
Each element carries its own unit, so one vector can mix coordinate systems, and
a grob can even use different units on its x and y axes.

The units you reach for most:

- `"npc"` (the default): normalised parent coordinates, `0` at bottom/left and
  `1` at top/right of the current viewport.
- `"native"`: the enclosing viewport's `xscale` / `yscale`, so data values map
  directly. This is what you use for plotted data.
- `"mm"`, `"cm"`, `"in"`, `"pt"`: absolute physical lengths that keep their size
  regardless of the viewport.

A bare number is interpreted in the grob's default units (usually `"npc"`), so
`x = 0.5` and `x = vl_unit(0.5, "npc")` are the same thing.

```{r}
#| label: units
vl_unit(1:3, "native")
vl_unit(c(0.5, 1), c("npc", "in"))
```

`"native"` units need a viewport with scales to resolve against. Set `xscale`
and `yscale` when you push:

```{r}
#| label: native
#| fig.width: 5
#| fig.height: 3
vl_scene(5, 3, bg = "white") |>
  push(vl_viewport(
    width = 0.86, height = 0.82,
    xscale = c(0, 10), yscale = c(-5, 25)
  )) |>
  draw(rect_grob(gp = vl_gpar(fill = "grey97", col = "grey70"))) |>
  draw(lines_grob(
    x = vl_unit(0:10, "native"),
    y = vl_unit((0:10) * 2, "native"),
    gp = vl_gpar(col = "steelblue", lwd = 2)
  )) |>
  pop()
```

Absolute and relative units compose within a viewport, and font- or
string-relative units (`"char"`, `"line"`, `"strwidth"`) resolve to millimetres
at construction. Mixing a relative and an absolute unit in a single arithmetic
expression (say `vl_unit(1, "npc") - vl_unit(2, "mm")`) is deferred and reported
rather than silently guessed, so an ambiguous offset fails loudly instead of
drawing in the wrong place.

## The paint model

Any `fill` in `vl_gpar()` can be more than a flat colour. The same three paint
types work identically on the raster, SVG, and PDF backends (with the documented
exception that the PDF backend does not yet rasterise patterns).

### Gradients

`linear_gradient()` and `radial_gradient()` interpolate between colour stops.
Their geometry is given in a coordinate system (`"npc"` by default) and is
resolved against the viewport at draw time, so a gradient transforms with its
grob just like the outline does.

```{r}
#| label: gradients
#| fig.width: 6
#| fig.height: 2.2
vl_scene(6, 2.2, bg = "white") |>
  push(vl_viewport(x = 0.28, width = 0.44)) |>
  draw(rect_grob(
    width = 0.8, height = 0.7,
    gp = vl_gpar(fill = linear_gradient(c("#1b2a4a", "#3a7bd5")), col = NA)
  )) |>
  pop() |>
  push(vl_viewport(x = 0.72, width = 0.44)) |>
  draw(circle_grob(
    r = 0.34,
    gp = vl_gpar(fill = radial_gradient(c("#f6d365", "#fda085")), col = NA)
  )) |>
  pop()
```

Stops blend in sRGB by default. Blending two distant hues there passes through a
muddy, over-dark middle — a blue→yellow ramp dips through grey. Set
`interpolation = "oklab"` to blend in the perceptually-uniform Oklab space
instead, so the ramp stays even and vivid. The same option works on every backend
(the stops are pre-sampled in Oklab into ordinary sRGB stops), and the default
`"srgb"` is unchanged.

```{r}
#| label: gradient-interpolation
#| fig.width: 6
#| fig.height: 1.6
bar <- function(x, interp) {
  rect_grob(
    x = x, width = 0.46, height = 0.7,
    gp = vl_gpar(col = NA, fill = linear_gradient(c("blue", "yellow"), interpolation = interp))
  )
}
vl_scene(6, 1.6, bg = "white") |>
  draw(bar(0.27, "srgb")) |> # sRGB: grey dead-zone in the middle
  draw(bar(0.73, "oklab")) # Oklab: clean cyan/green transition
```

### Patterns

`vl_pattern()` fills a shape by tiling a grob (or a list of grobs). The tile is
authored in the unit square and repeated across a cell whose size you choose.

```{r}
#| label: pattern
#| fig.width: 4
#| fig.height: 2.4
tile <- list(
  rect_grob(gp = vl_gpar(fill = "#ecf0f1", col = NA)),
  circle_grob(r = 0.32, gp = vl_gpar(fill = "#e74c3c", col = NA))
)

vl_scene(4, 2.4, bg = "white") |>
  draw(rect_grob(
    width = 0.84, height = 0.84,
    gp = vl_gpar(fill = vl_pattern(tile, width = 0.18, height = 0.3), col = NA)
  ))
```

### Masks and group opacity

A mask is a grob whose coverage modulates the visibility of a viewport's
contents. Wrap it with `as_mask()` and pass it to `vl_viewport(mask = ...)`. Here a
linear gradient is clipped to a circular alpha mask.

```{r}
#| label: mask
#| fig.width: 4
#| fig.height: 2.4
vl_scene(4, 2.4, bg = "white") |>
  push(vl_viewport(
    mask = as_mask(circle_grob(r = 0.42, gp = vl_gpar(fill = "white", col = NA)))
  )) |>
  draw(rect_grob(gp = vl_gpar(fill = linear_gradient(c("#7f53ac", "#647dee")), col = NA))) |>
  pop()
```

Related to masks is **group opacity**. Setting `vl_viewport(alpha = ...)`
composites the viewport's contents as a single isolated layer at that opacity,
so overlapping elements do not accumulate the way per-element `vl_gpar(alpha = )`
would. That distinction (compositing a group versus fading each mark) is exactly
the kind of control a grammar layer needs from its backend.

## Recap

- A scene is a retained tree of nested viewports and grobs, built with
  `push()` / `draw()` / `pop()`.
- `vl_unit()` vectors express geometry; `"npc"` is relative to the viewport,
  `"native"` follows the data scales, and `"mm"` and friends are absolute.
- `vl_gpar(fill = )` accepts gradients and patterns, and `vl_viewport()` accepts masks,
  group opacity, and blend modes, all consistent across backends.

Next, see `vignette("retained-mode")` for what the retained tree lets you do
after it is built.
```
