---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3,
  dpi = 96,
  dev = "png"
)
```

vellum is a low-level graphics framework for R, in the spirit of grid. You
describe a scene with a small declarative API and render it to PNG, SVG, or PDF.

The property most of the rest follows from is that vellum measures text and
solves layout itself, in process, instead of asking whichever device happens to
be current. So a scene has a size before there is a page, layout is solved once
rather than once per output format, and — the part with no counterpart in grid —
a finished scene can report where it put every element (`scene_model()`) and
answer what lies under a given point (`hit_test()`). That read-back is what an
interactive host, an accessibility layer, or a test suite binds to.

It is the foundation layer a grammar of graphics builds on, the way grid
underlies ggplot2. It is not a plotting package itself, so there are no scales,
stats, geoms, or facets here. What it gives you is the drawing substrate: units,
viewports, grobs, layout, a deterministic renderer, and a scene you can query
afterwards.

```{r}
#| label: setup
library(vellum)
```

## A first scene

A scene starts with `vl_scene()`, which fixes the page size (in inches by
default) and background. You then add content with a pipeline of `draw()` calls
and show the result. Auto-printing a scene displays it, so the last line of a
chunk is enough.

```{r}
#| label: hello
vl_scene(width = 6, height = 3, bg = "white") |>
  draw(rect_grob(
    width = 0.94, height = 0.82,
    gp = vl_gpar(fill = linear_gradient(c("#1b2a4a", "#3a7bd5")), col = NA)
  )) |>
  draw(circle_grob(
    x = 0.16, y = 0.5, r = 0.28,
    gp = vl_gpar(fill = "#f7c948", col = NA)
  )) |>
  draw(text_grob(
    "vellum", x = 0.62, y = 0.5,
    gp = vl_gpar(fontsize = 48, col = "white", fontface = "bold")
  ))
```

Three ideas do most of the work here.

**Grobs** are the drawable primitives: `rect_grob()`, `circle_grob()`,
`text_grob()`, and a couple of dozen more (see `?grob`). Each is an immutable
value object; building one draws nothing on its own.

**`vl_gpar()`** carries the graphical parameters (fill, stroke colour, line width,
font, opacity) attached to a grob. A `fill` can be a plain colour or a gradient,
as above.

**Coordinates default to `"npc"`**: normalised parent coordinates, where `(0, 0)`
is the bottom-left of the region and `(1, 1)` the top-right. So `x = 0.16` sits
near the left edge and `y = 0.5` is vertically centred.

## Building a scene: push, draw, pop

A scene is a tree of viewports and grobs, built functionally. `push()` descends
into a new `vl_viewport()`, `draw()` adds a grob at the current level, and `pop()`
ascends. A viewport is a rectangular region that establishes its own coordinate
systems, so once you push one with an `xscale` and `yscale`, `"native"` units
map data values straight onto the region.

```{r}
#| label: scatter
#| fig.width: 5
#| fig.height: 3.2
set.seed(1)
x <- runif(60, 0, 10)
y <- 1.8 * x + rnorm(60, 0, 4)

vl_scene(5, 3.2, bg = "white") |>
  push(vl_viewport(
    x = 0.57, y = 0.57, width = 0.82, height = 0.82,
    xscale = c(0, 10), yscale = range(pretty(y))
  )) |>
  draw(rect_grob(gp = vl_gpar(fill = "#f4f6f8", col = "#cfd8dc"))) |>
  draw(points_grob(
    vl_unit(x, "native"), vl_unit(y, "native"),
    size = vl_unit(3.2, "mm"),
    gp = vl_gpar(fill = "#3a7bd5", col = "#1b2a4a", lwd = 1)
  )) |>
  pop()
```

The panel rectangle is drawn in `"npc"` (it fills its viewport), while the points
are placed in `"native"` units, so their positions follow the data scales. A
single `vl_unit()` vector can even mix coordinate systems per axis. See
`vignette("scene-and-paint")` for the full picture of units, viewports, and the
paint model.

## Rendering to a file

Displaying a scene is convenient while exploring. To write output, call
`render()`; it picks the backend from the file extension.

```{r}
#| label: render
#| eval: false
s <- vl_scene(4, 3) |>
  draw(circle_grob(r = 0.3, gp = vl_gpar(fill = "tomato", col = NA)))

render(s, "out.png") # raster (tiny-skia)
render(s, "out.svg") # vector (hand-rolled SVG)
render(s, "out.pdf") # vector (krilla)
```

The same scene value renders to all three formats with consistent geometry, and
raster and PDF output is byte-stable, so it is reproducible and
snapshot-testable.

`render()` takes two arguments worth knowing about. `scale` renders at a
multiple of the device pixels while keeping the same physical size — the retina
idiom — so nothing about the layout or the relative size of text changes. `cvd`
simulates a colour-vision deficiency, which is covered in
`vignette("render-quality")`.

```{r}
#| label: render-args
#| eval: false
render(s, "out@2x.png", scale = 2) # twice the pixels, same physical size
render(s, "check.png", cvd = "deuteranopia") # as a red/green-blind reader sees it
```

You do not have to go through a file. Every format is also available in memory,
which is what a web handler or an HTML widget wants:

```{r}
#| label: in-memory
#| eval: false
scene_png(s) # raw vector: the encoded PNG
scene_pdf(s) # raw vector: the encoded PDF
scene_svg(s) # character: the SVG document
scene_raster(s) # integer array: the pixels themselves
```

## Where to go next

- `vignette("scene-and-paint")`: the scene graph, units and viewports, and the
  paint model (gradients, patterns, masks).
- `vignette("retained-mode")`: read back per-element geometry with
  `scene_model()`, pick a node under a point with `hit_test()`, and edit nodes by
  name.
- `vignette("inspecting-scenes")`: `vl_lint()` to catch likely mistakes,
  `scene_stats()` for coverage and crowding, `profile_render()` for where the
  time goes.
- `vignette("render-quality")`: colour-vision simulation, crisp gridlines, and
  group blur/shadow effects.
- `vignette("typography")`: text halos and OpenType features.
- `vignette("coming-from-grid")`: a translation guide if you already know grid.
- `vignette("grid-interop")`: render existing grid, ggplot2, and lattice output
  through the vellum backend.
```
